Stage-Splitting a Modal Language

### 2014-10-20 paper/example.tex
first pass on example.tex

@@

\ur{$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.}

@@

\ur{I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.}

@@

Knoblock and Ruf~\cite{KR} proposed  the {\em data specialization} technique

limitations of

Prior work on

>>>>

\subsection{Comments, suggestions, and ``todo''}
I rewrote this section.  There is more work to be done.

\begin{itemize}
\item 
I find it hard to distinguish between blue and black on printed paper.

\item
$\pause$ should be called ``promote'' or something like that. I
  believe this is what Taha calls cross stage persistence.  THat work should
  be cited.

\item
I don't like fuzzy words like hygenic that dont mean anything.
  What is the point of using them?  A paper ought to be written for
  the reader not as a note to self.
\end{itemize}

@@

@@

@@

Knoblock and Ruf~\cite{KR} proposed the {\em data specialization}
technique to overcome some of these limitations of program
specialization.  The idea behind this technique is to take a staged
program and split it into a stage-1 and a stage-2 program without
knowledge of any of the arguments.  Unlike program specialization,
data specialization does not require knowledge of the first-stage
arguments to produce fast code for the second stage.  Instead it yield
a first stage program that maps the first stage arguments to an
intermediate or {\em boundary data structure}, which is used by the
second stage program along with the second stage arguments to produce
the final results.  Knoblock and Ruf's observation is that the
information required by the second stage can be represented solely as
a data structure, making it unnecessary to generate and compile or
interpret code dynamically at run time.  Data specialization can
therefore be applied statically at compile time, leading to
drastically improved run-time performance without any significant
run-time overheads.

By applying data specialization to our quickselect example, we would
wish to derive two functions, \texttt{quickselect1} and
\texttt{quickselect2}, that would be semantically equivalent to
quickselect in the following sense:
\\
\texttt{quickselect l k = quickselect2 (quickselect1 l) k}.
\\
%
Since \texttt{quickselect1} requires only on a first stage
argument~\texttt{l}, we can run it once to produce a boundary data
structure that can be used for all subsequent queries on~\texttt{l},
e.g.,  the code consisting of $m$ calls 
\begin{lstlisting}
quickselect l 1
quickselect l 2
 @$\ldots$@ 
quickselect l @$m$@
\end{lstlisting}
%
can be replaced by 
%
\begin{lstlisting}
b = quickselect1 l
quickselect2 b 1
quickselect b 2
@$\ldots$@ 
quickselect b @$m$@.
\end{lstlisting}

The benefit of this ``optimization'' can be dramatic.  For example, in
principle at least, the first stage function \texttt{quickselect1}
could produce a balanced binary search tree holding the elements
of~\texttt{l} and the second stage function \texttt{quickselect2}
could use this balanced binary search tree and the given rank value to
perform a logarithmic-time search in the tree to return the desired
result.  Assuming that the list~\texttt{l} contains $n$ elements, this
optimization would change the asymptotic complexity from expected
(randomized) $\Theta(n \cdot m)$ to $\Theta(n\log{n} \cdot m
\cdot\log{n})$, which for any $m \approx n$ reduces the complexity
from $\Theta (n^2)$ to $\Theta(n\log{n})$---a near linear time
improvement.  Note that the desired optimized code shows above is
intellectually more sophisticated than the code that we have started
with: the optimized code is able to create a data structure, a
balanced binary tree augmented with indexing information, and use a
binary search technique over this tree to compute the result
asymptotically more efficiently.  In fact, based our teaching
experience, we can imagine this kind of problem to be a moderately
difficult exam question in an undergraduate algorithms class, as it
not only requires understanding of data structures such as binary
search trees but also requires modifying them to augment with indexing
information to support rank-based search.

It is thus perhaps no surprise that the prior art in data
specialization is not able to perform such ``optimizations''.
Knoblock and Ruf's paper takes a very important step by formulating
the problem and proposing a solution.  But their work applies to
program fragments that don't involve sophisticated data types such as
as sums and recursive types, and don't involve recursion. For example,
their work would not apply to our quickselect example and similar ones
involving recursion and/or  recursive types.

In this paper, we show that it is possible to generalize data
specialization to a fully general language with recursion and
recursive types.  The key ideas behind our approach are: ...

In what follows we present a relatively mechanical application of our
techniques to our running example.

\ur{I made up to here.  What follows should be shortened and made to
  the point. Some time later...}

### 2014-10-20 paper/paper.tex
added remarks
first pass on example.tex

\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}

@@

\input {macros}

\definecolor{stageOneColor}{RGB}{120,30,0}
\definecolor{stageTwoColor}{RGB}{0,0,127}
\begin{document}

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
        escapeinside={"*}{*"},
        moredelim=*[is][\color{stageOneColor}]{1`}{`},
        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

@@

\input{intro}

@@

%\vfill\eject\pagebreak
\input{splitting}
\input{algoexamples}
\input{relatedwork}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
%\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}
\usepackage{verbatim}

@@

\input{macros}
\input{macros-ref}
\input{remarks}

\definecolor{stageOneColor}{RGB}{120,30,0}
\definecolor{stageTwoColor}{RGB}{0,0,127}

\begin{document}

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
%        escapeinside={"*}{*"},
        escapeinside={@}{@},
        moredelim=*[is][\color{stageOneColor}]{1`}{`},
        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

@@

%% \input{intro}

@@

%% %\vfill\eject\pagebreak
%%\input{splitting}
\input{algoexamples}
\input{relatedwork}
\input{umut}

### 2014-10-20 paper/umut.tex
first pass on example.tex

The following definitions are adapted from the partial evaluation
book, the cited papers, and Malmkjaer thesis.  For simplicity and
uniformity, we assume that data and all programs are drawn from the
same set and that ill-behaved programs and non-terminating programs
all return $bottom$.

@@

@@

\subsection{Techniques}

>>>>

\paragraph{Defitions and relationships.}
The following definitions are adapted from the partial evaluation
book, the cited papers, and Malmkjaer thesis.  For simplicity and
uniformity, we assume that data and all programs are drawn from the
same set and that ill-behaved programs and non-terminating programs
all return $bottom$.

@@

THIS IS IMPORTANT. EXPAND ON THIS.  Knoblock and Ruf's data
specialization is similar but statically splits the code. This is
important because it is not feasible to perform splitting at run time
as can be possible according to the definition of data splitting. This
is similar to ours.  Their approach, however, seems to assume that the
specialized code is a single non-recursive procedure.

@@

\paragraph{Techniques}
